Accelerating parameter inference with graphics processing units

### 2019-03-20 paper/referee/response-v0.tex

In a future version of the code, described in the second paragraph of our conclusions, we intend to perform all Monte
Carlo random number generation on the GPU.  In this configuration we will not have significant data transfer on and off
the board.

\item * Relatedly, how much memory is required to store these matrices?
Since GPUs have smaller RAM sizes, is any special care or tuning
required to get the problem to fit onto the device? Later on the
authors use the ell=2 modes as example -- since the size of the
matrices grows with the number of modes, does the code need to account
for this in any way?

\noindent \textbf{Response}: While the memory footprint of our operations loosely grows linearly with the number of
modes involved, we have noticed no significant limitations on our memory footprint.  We have colleagues who use models with modes
$l \le 8$ without reporting memory footprint problems.

The memory footprint of the initially transferred data is dominated by the $Q_{k,lm}$ time arrays is no larger than $16 kHz
\times 0.15  \lesssim 2500$ floating point numbers per mode per detector.  Even with several hundred modes, these are
quite small.
Potentially more problematic are our intermediate data products, notably the matrix of time-shifted arrays $Q$ built
from $Q_{k,lm}(\lambda,\tau)$ .  Our code has been carefully
organized to loop over detectors so the memory footprint of $Q$ is only (modes)$\times$ (extrinsic)$\times$ (time),
which (because we perform evaluations in batches of  $10^4$ extrinsic parameters at a time) is roughly $10^4$ times the
footprint of all the underlying $Q_{k,lm}(\lambda,\tau)$ timeseries.  So many modes can
fit well within a typical 4Gb GPU memory space, even allowing for substantial overhead.

@@

\noindent \textbf{Response}:  We perform our timing estimates on individual runs on busy nodes and clusters on which we
have shared access, not
dedicated hardware.  We anticipate small timing discrepancies from run to run.

\editremark{GO BACK AND CHECK THESE NUMBERS}

\item * The authors focus on computing posteriors of intrinsic parameters.
But for low-latency studies, the extrinsic parameters are arguably
more important as they can be used to direct optical telescopes. Can
the authors please comment on what changes (if any) are needed to do
low-latency studies for producing, for example, sky maps? Is this a
"trivial" change to the code or will it require more significant
re-working? I ask because fully Bayesian low-latency sky maps would be
very useful, so if this code can currently do that then it would be
worth mentioning it more prominently. Otherwise, it would be worth
discussing what additional work will be needed to realize this
promise.

\noindent \textbf{Response}:  We have added a brief comment in our conclusions to emphasize that extrinsic information
has previously been produced with a similar pipeline to the one benchmarked  here,  and that information will be
made rapidly available in an automated mode by this pipeline very soon.

We point out however that our inference strategy does not presently account for the effect of calibration error, which can have a
notable impact on sky localization: it's not immediately clear our strategy would necessarily provide a dramatic change
over the Bayestar skymaps, except insofar as we perform inference with a fully precessing signal model (and/or a model
containing higher modes).     By contrast, LI inference can account for the impact of calibration error on sky localization.

@@

\item * Fig 1's caption: what does "on the evaluation points" mean? What are
the evaluation points -- maybe this could be explained in a bit more
detail in the main body of the paper.

\noindent \textbf{Response}: We modified the caption to define the set whose mean and variance is being evaluated.

@@

\item * Figure 1: is the quantity being plotted in the lower left corner
actually the mean of $-log(L_{marg})$? (not the minus sign).

\noindent \textbf{Response}: No, the sign is correct.  Due to the normalization convention adopted (see response \# 2), the peak value
of our likelihood (ratio) is greater than zero.

@@

\item * Figure 2: same questions I have for figure 1 apply here.

\noindent \textbf{Response}: The sign is correct.

\item * Page 7: "Our initial grid consists of 5000 points, spread uniformly
across a 4-dimensional hypercube"... do the authors mean to say $9^4 =
6561$ points?

\noindent \textbf{Response}: We modified the text to add the word ``approximately'' before uniformly.

In practice, our grid layout code returns precisely the number of points requested.  It attempts to choose the sizes of
each edge of the hypercube to best conform to but be strictly larger than the target size, and then truncates the set of
candidate points.

@@

\item * To give the timing numbers (ie the walltime) a bit more context,
please consider adding the time it would take to do a similar analysis
(same recovery model, set of modes, and similar CPU hardware) with
LALInference. Even rough estimates would be very helpful towards
understanding the full impact of the GPU-accelerated rapidPE code. I
feel like this is one of the most important points, as it is the key
to the impact of this work.

\noindent \textbf{Response}:  In the conclusions, we \editremark{will} have added some anecdotal apples-to-apples timing
information on runs of practical interest.

That said, a direct apples-to-apples comparison between RIFT and LI using \emph{identical} waveforms and settings to
what we benchmark here can be grossly unfair to LI, even though it is an apples-to-apples comparison.  LI developers
have highly prioritized the use of fast approximations, and reduced-order-quadratures that further accelerate the use of
these approximations.     To the extent these approximations don't impact results, LI can often perform comparable
calculations much more quickly than a configuration which used apples-to-apples settings.

Of course, these approximations are approximations.  Existing approximations  have been demonstrated by example to have limitations /
introduce biases that, if left unchecked, would be increasingly significant as observaitons accumulate (e.g., Lange et
al 2018).  RIFT's rapid performance  and low cost can enable systematics investigations which would otherwise be inaccessible.

@@

\item * The fonts in some of the figures, and especially figure 3, are small
and in almost unreadable. Please improve the readability of these
figures.

\noindent \textbf{Response}:  \editremark{fix me: rcparams}

\item * I would appreciate it if the authors could explain to me why the
evidence converges much more quickly in figure 3 as compared to
figures 1 and 2. Or said in a different way, way are the shapes of
evidence vs iteration so different. They may optionally wish to
comment on this in the paper too.

\noindent \textbf{Response}:  The convergence rate of the evidence depends on the system studied and, critically, the
initial grid: if we have a very well-chosen initial starting grid, the fit  converges rapidly, potentially after just
one iteration.  The configuration used
in Fig 3 has a different dimensionality (4d, not 2d) and a larger starting  grid than our lower-dimensionality investigation.

Qualitatively speaking and based on our large but anecdotoally-built experience, we anticipate that  due to strong degeneracy in
spin -- only one spin degree of freedom is well-constrained -- the seemingly-sparse grid in $\mc$ and $\delta$ is very
informative because several redundant configurations with similar $\chi_{\rm eff}$ but different $\chi_{i,z}$ inform our
fit.

A new  student here has begun systematically investigating parameters that impact the convergence rate, using controlled
circumstances. The results of these investigations are too preliminary to report here.

@@

\item * "Reducing the sampling rate by a factor s will the cost of all
operations with timeseries – they are shorter."

\noindent \textbf{Response}:added ``reduce''

\item * "or by marginally more conservative convergence thresholds. on N
eval or N it"

\noindent \textbf{Response}: Changed ``. on $N_{eval}$'' to ``$n_{\rm eff}$''

@@

\item *"Fortunately, the number of observationally significant and
accessible dimensions is often substantially less than the a prior
necessary dimensionality."

\noindent \textbf{Response}: removed ``the'' and ``necessary'' in the last few words
